Causal Impact of Masks, Policies, Behavior on Early Covid-19 Pandemic in the U.S.

### 2020-09-06 tex/reply-to-referees.tex

We believe that it is better to use the number of cases and deaths rather than time since start of the outbreak as a  general framework to analyze the spread of COVID-19. This is especially true if we analyze the second wave of COVID-19, where time since start of the first outbreak may not be so relevant, and identifying the start of the second outbreak may be difficult.

\item \textit{`` Figure 8 - I do like seeing the data, but it also just raises questions, as you say, confounders + many other things at play.... I wondered if there would be a way to create a visual that captured a bit more of this, and pinpointed the anomalies which are where the inferential power lies. ''}

This is an interesting suggestion. We thought about the possibility of providing a version of Figure 8 (now, Figure 10) based on the regression residuals after taking out the effect of confounders etc but, eventually, we decided not to present such results due to the space constraint.

For your reference, we have constructed a version of  figures based on the residuals of regressing the case or death growth rates on state characteristics, month dummies,   their interactions,  the log of lagged cases/deaths, and the growth rate of lagged cases/deaths, which illustrates that states with a mask mandate tend to have lower case and death growths even after controlling for confounders and information variables as below.   % and then computing their average value for states with or without a mask mandate for each date.

\begin{figure}[ht]
  \caption{Average  residuals of regressing case and death growth on confounders among states with or without  mask mandates \label{fig:masks-res}}\vspace{0.2cm}
 \begin{minipage}{\linewidth}
    \centering
    \begin{tabular}{cc}
      \includegraphics[width=0.5\textwidth]{tables_and_figures/pmaskbus-cases-res-14}
      &
        \includegraphics[width=0.5\textwidth]{tables_and_figures/pmaskbus-deaths-res-21}
    \end{tabular}
    \begin{flushleft}
      \scriptsize In these figures, red points are the regression residuals of case or death
      growth rate on confounders and information variables in states without a mask mandate. Blue points are
      states with a mask mandate 14 (21 for deaths) days prior. The
      red line is the average across states without a mask mandate 14
      (21 for deaths) days earlier. The blue line is the average
      across states with a mask mandate 14 (21 for deaths) earlier.
    \end{flushleft}
  \end{minipage}
\end{figure}

### 2020-09-06 cases_and_policies/rmd/regressionsWithPlots.Rmd
update

# use residuals after controlling for confounders 
wv <- c(svars,mvars.c,msvars.c,"month3") 
fmla <- as.formula(paste("dlogdc", "~",  paste(wv, collapse=" + "), sep=" "))
g1 <- predict(lm(fmla, data=df))
df$dlogdc.res <- df[,yv] - g1    
fig <- pfigure(df, "pmaskbus", stlab=-1,outcome="dlogdc.res",L=L.c) + ggtitle(TeX(relabel("dlogdc given pmaskbus"))) + ylab(TeX(relabel("residual of dlogdc"))) + 
  ylim(c(-1.8,0.5))  
p <- "pmaskbus"
figdev(sprintf("%s/tex/tables_and_figures/%s-cases-res-%s.pdf", rootdir, p, L.c), width=4, height=3)
print(fig)
dev.off()
fig

wv <- c(svars,mvars.d,msvars.d,"month3") 
fmla <- as.formula(paste("dlogdd", "~",  paste(wv, collapse=" + "), sep=" "))
g1 <- predict(lm(fmla, data=df))
df$dlogdd.res <- df[,yv] - g1  
fig <- pfigure(df, "pmaskbus", stlab=-1,outcome="dlogdd.res",L=L.d) + ggtitle(TeX(relabel("dlogdd given pmaskbus"))) + ylab(TeX(relabel("residual of dlogdd")))+ 
  ylim(c(-0.9,0.5))  
p <- "pmaskbus"
figdev(sprintf("%s/tex/tables_and_figures/%s-deaths-res-%s.pdf", rootdir, p, L.d), width=4, height=3)
print(fig)
dev.off()
fig

>>>>

# use residuals after controlling for confounders

sdf <- subset(df, df$date>=as.Date("2020-03-07")) #+L)
sdf$one <- 1
wv <- c(svars,mvars.c,msvars.c,"month3","one")
wv <- c(svars,mvars.d,msvars.d,"month3","one","dlogdc.L14","logdc.L14")#,"dlogdc.national.L14","logdc.national.L14")
fmla <- as.formula(paste("dlogdc", "~",  paste(wv, collapse=" + "), sep=" "))
g1 <- predict(lm(fmla, data=sdf))
sdf$dlogdc.res <- sdf$dlogdc - g1
mean(sdf$dlogdc.res)
fig <- pfigure(sdf, "pmaskbus", stlab=-1,outcome="dlogdc.res",L=L.c) + ggtitle(TeX(relabel("dlogdc given pmaskbus"))) + ylab(TeX(relabel("residual of dlogdc"))) + 
 #   xlim(c(as.Date("2020-04-01"), max(df$date))) +
  ylim(c(-0.8,0.3))  
p <- "pmaskbus"
figdev(sprintf("%s/tex/tables_and_figures/%s-cases-res-%s.pdf", rootdir, p, L.c), width=4, height=3)
print(fig)
dev.off()
fig

wv <- c(svars,mvars.d,msvars.d,"month3","one")
wv <- c(svars,mvars.d,msvars.d,"month3","one","dlogdd.L21","logdd.L21")#,"dlogdd.national.L21","logdd.national.L21")
fmla <- as.formula(paste("dlogdd", "~",  paste(wv, collapse=" + "), sep=" "))
g1 <- predict(lm(fmla, data=sdf))
sdf$dlogdd.res <- sdf$dlogdd - g1  
fig <- pfigure(sdf, "pmaskbus", stlab=-1,outcome="dlogdd.res",L=L.d) + ggtitle(TeX(relabel("dlogdd given pmaskbus"))) + ylab(TeX(relabel("residual of dlogdd")))+ 
 #   xlim(c(as.Date("2020-04-01"), max(df$date))) +
  ylim(c(-0.8,0.3))  
p <- "pmaskbus"
figdev(sprintf("%s/tex/tables_and_figures/%s-deaths-res-%s.pdf", rootdir, p, L.d), width=4, height=3)
print(fig)
dev.off()
fig

### 2020-09-06 tex/covidEffectOfNPI-revision.tex
update

A growing number of other papers have examined the link between non-pharmaceutical interventions and Covid-19  cases.\footnote{We refer the reader to \cite{avery2020} for a comprehensive review of  a larger body of work researching Covid-19; here we focus on few quintessential comparisons on our work with other works that we are aware of.} \cite{hsiang2020} estimate the effect of policies on the growth rate of cases using
data from the United States, China, Iran, Italy, France, and South
Korea. In the United States, they find that the combined effect of all policies they consider on the
growth rate is $-0.347$ $(0.061)$. \cite{courtemanche2020} use US county level data to analyze the effect
of interventions on case growth rates. They find that the combination
of policies they study reduced growth rates by 9.1 percentage points
16-20 days after implementation, out of which 5.9 percentage points are attributed to shelter in place orders. Both \cite{hsiang2020} and \cite{courtemanche2020} adopted a reduced-form approach to estimate the total policy effect on case growth without using any social distancing behavior measures. In contrast, our study highlights the role of behavioral response to policies and information.

@@

Given the lack of experimental evidence on the effect of masks in the context of COVID-19, conducting observational studies  is useful and important. To the best of our knowledge, our paper is the first  empirical study that shows the effectiveness of  mask mandates on reducing the spread of Covid-19 by analyzing the US state-level data. This finding corroborates and is complementary to the medical observational evidence in \cite{howard2020}. Analyzing mitigation measures in New York, Wuhan, and Italy, \cite{zhangr2020} conclude that mandatory face coverings substantially reduced infections. \cite{abaluck2020}  find that the growth rates of cases and of deaths in countries with pre-existing norms that sick people should wear masks are lower by 8 to 10\% than those rates in countries with no pre-existing mask norms.\footnote{\cite{miyazawa2020} find that  country's COVID-19 death rates are negatively associated with mask wearing rates.} Our finding is also independently corroborated by a completely different causal methodology based on synthetic control
using German data in \cite{Mitze2020}.\footnote{Our study was first released in ArXiv on May 28, 2020 whereas
 \cite{Mitze2020} was released at SSRN on June 8, 2020.}

>>>>

A growing number of other papers have examined the link between non-pharmaceutical interventions and Covid-19  cases.\footnote{We refer the reader to \cite{avery2020} for a comprehensive review of  a larger body of work researching Covid-19; here we focus on few quintessential comparisons on our work with other works that we are aware of.} \cite{hsiang2020} estimate the effect of policies on the growth rate of cases using
data from the United States, China, Iran, Italy, France, and South
Korea. In the United States, they find that the combined effect of all policies they consider on the
growth rate is $-0.347$ $(0.061)$. \cite{courtemanche2020} use US county level data to analyze the effect
of interventions on case growth rates. They find that the combination
of policies they study reduced growth rates by 9.1 percentage points
16-20 days after implementation, out of which 5.9 percentage points are attributed to shelter in place orders. Both \cite{hsiang2020} and \cite{courtemanche2020} adopted a reduced-form approach to estimate the total policy effect on case growth without using any social distancing behavior measures.\footnote{Using
a synthetic control approach, \cite{cho2020} finds that  the cases would have been lower by 75 percent  had Sweden adopted stricter lockdown policies.}

@@

Given the lack of experimental evidence on the effect of masks in the context of COVID-19, conducting observational studies  is useful and important. To the best of our knowledge, our paper is the first  empirical study that shows the effectiveness of  mask mandates on reducing the spread of Covid-19 by analyzing the US state-level data. This finding corroborates and is complementary to the medical observational evidence in \cite{howard2020}. Analyzing mitigation measures in New York, Wuhan, and Italy, \cite{zhangr2020} conclude that mandatory face coverings substantially reduced infections. \cite{abaluck2020}  find that the growth rates of cases and of deaths in countries with pre-existing norms that sick people should wear masks are lower by 8 to 10\% than those rates in countries with no pre-existing mask norms.\footnote{\cite{miyazawa2020} find that  country's COVID-19 death rates are negatively associated with mask wearing rates.} The Institute for Health Metrics and Evaluation at the University of Washington  assesses that, if 95\% of the people in the US were to start wearing masks from early August of 2020, 66,000 lives would be saved by December 2020 \citep{IHME2020}.
Our finding is also independently corroborated by a completely different causal methodology based on synthetic control
using German data in \cite{Mitze2020}.\footnote{Our study was first released in ArXiv on May 28, 2020 whereas
 \cite{Mitze2020} was released at SSRN on June 8, 2020. }
